\large \texttt{insectDisease}: programmatic access to the \textit{Ecological Database of the World's Insect Pathogens}

### 2021-08-31 manuscript/manuscript.tex
updates to vignette

\paragraph*{}
Data are also available on the insect host species themselves (e.g., \texttt{?hosts} or \texttt{data(hosts)}). These data contain some information on Canadian province where the host is found (\texttt{ProvinceI} column), what it eats (\texttt{Food} column), and what type of habitat it is found in (\texttt{Habitat} column). Additionally, a column on host insect pest status is present, offering the opportunity to explore study effort and pathogen specificity dependent on the pest status of the insect host.

@@

The initial data structure was maintained from the original raw data files provided by David Onstad, principal maintainer of the EDWIP data resource \citep{onstad1997}. This includes files such as \texttt{new\_assoc}, as this was likely a test file containing pathogen species such as "wormy thing". We clean and augment this existing data source programmatically, with much of this code in the \texttt{insectDisease} package \texttt{vignette}.

\paragraph*{}
First, we resolve host and pathogen names using the R package \texttt{taxize}, using NCBI taxonomic backbone \citep{chamberlain2013}. While this seems like a small change, it both standardizes host and pathogen nomenclature, and catches any taxonomic changes that have occurred in the past couple decades. This includes the consideration of microsporidian parasites as fungi, not protozoans, a change affecting a large set of records in the EDWIP data. Second, we provide a vignette which combines the different host-pathogen data together to form an informative set of host-pathogen associations, including both both known associations and failed experimental infections, as delineated by the \texttt{interaction} column in the resulting \texttt{edwip.csv} file. Finally, processed data by the vignette is output in \texttt{.csv} format in the vignettes folder, but all data stored as R data objects (\texttt{.rda}) are also converted to csv and placed in the \texttt{csv} folder, allowing non-R users to still access the data in a stable format.

>>>>

\paragraph*{}
Data are also available on the insect host species themselves (e.g., \texttt{data(hosts)}). These data contain some information on Canadian province where the host is found (\texttt{ProvinceI} column), what it eats (\texttt{Food} column), and what type of habitat it is found in (\texttt{Habitat} column). Additionally, a column on host insect pest status is present, offering the opportunity to explore study effort and pathogen specificity dependent on the pest status of the insect host.

@@

We attempted to maintain as much of the original data structure from the raw data files provided by David Onstad, principal maintainer of the EDWIP data resource \citep{onstad1997}. This includes files such as \texttt{new\_assoc}, as this was likely a test file containing pathogen species such as "wormy thing". We provide comments in the documentation about which data sources are perhaps more trustworthy. We clean and augment this existing data source programmatically, with much of this code in the \texttt{insectDisease} package \texttt{vignette}.

\paragraph*{}
The first, and perhaps most important augmentation, is the resolution of host and pathogen taxonomic information. We achieved this by using the R package \texttt{taxize}, specifically the NCBI taxonomic backbone \citep{chamberlain2013}. Cached versions of host and pathogen taxonomic information are provided (\texttt{data(hostTaxonomy)} and \texttt{data(pathTaxonomy)}), and the $R$ code to generate these taxonomic backbones and clean the data are provided in the package vignette. This taxonomic backbone serves to both standardize host and pathogen nomenclature, while also correcting any taxonomic changes that have occurred in the past couple decades. This includes the consideration of microsporidian parasites as fungi, not protozoans, a change affecting a large set of records in the EDWIP data. Second, we provide a vignette which transparently shows the steps to clean and augment the data resource, as well as reproduce figures from this manuscript. Further, the vignette includes code which generates host-pathogen associations for all the data combined into a single data resource, simplifying the data down to a small number of important columns. Finally, processed we opt to store raw data in the \texttt{csv} folder, which contains all raw data files in .csv format. These data are also provided as .rda files in the \texttt{data} folder. This allows non-$R$ users to access the csv-formatted data easily, and ensures long-term stability of the data, as \texttt{csv} is a stable text file format.
